Teaching modeling in introductory statistics: A comparison of formula and tidyverse syntaxes

### 2023-01-13 PaperDraft.Rmd
next major version of paper

abstract: >
  This paper reports on a head-to-head comparison run in a pair of introductory statistics labs, one conducted fully in the formula syntax, the other in tidyverse. Analysis of incidental data from YouTube and RStudio Cloud show interesting distinctions. The formula section appeared to watch a larger proportion of pre-lab YouTube videos, but spend less time computing on RStudio Cloud. Conversely, the tidyverse section watched a smaller proportion of the videos and spent more time computing. Analysis of lab materials showed that tidyverse labs tended to be slightly longer in terms of lines in the provided RMarkdown materials, but not in minutes of the associated YouTube videos. The tidyverse labs exposed students to slightly more distinct R functions, but both labs relied on a quite small vocabulary of consistent functions, which can provide a starting point for instructors interested in teaching introductory statistics in R. Analysis of pre- and post-survey data show no differences between the two labs, so students appeared to have a positive experience regardless of section. This work provides additional evidence for instructors looking to choose between syntaxes for introductory statistics teaching. 
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    keep_tex: yes
    includes:
      in_header: header.tex
    citation_package: natbib
header-includes:
-  \usepackage{amsmath}
editor_options: 
  chunk_output_type: console
---

@@

```{r, cache = FALSE}
library(here)
library(tidyverse)
library(kableExtra)
library(lme4)
library(broom.mixed)
library(bookdown)
ggplot2::theme_set(ggplot2::theme_gray(base_size = 14))
```

@@

For example, the UC Berkeley Data 8 course uses Python, including elements of the commonly-used `matplotlib` and `numpy` libraries as well as a specialized library written to accompany the curriculum called `datascience` [@adhikarietal2021; @deneroetal2020]. The `datascience` library was designed to reduce complexity in the code. At the K-12 level, the language Pyret has been developed as a simplified version of Python to accompany the Bootstrap Data Science curriculum [@krishnamurthietal2020].

In R, the development of less-complex code for students has been under consideration for even longer [@pruimetal2011]. R offers non-standard evaluation, which allows package authors to create new 'syntax' for their packages [@morandatetal2012]. In human language, syntax is the set of rules for how words and sentences should be structured. If you use the wrong syntax in human language, people will hear there is something wrong with how you structured your speech or writing. However, because human understanding is flexible, the listener will probably still understand the general idea you were trying to convey. Syntax in programming languages is more formal-- it governs what code will execute, run, or compile correctly. Using the wrong syntax usually means failing to get a result from the program.

Typically, programming languages have only one valid syntax. For example, an aphorism about the language Python is "There should be one-- and preferably only one --obvious way to do it" [@peters2004]. But, non-standard evaluation in R has allowed there to be many obvious ways to do the same task. There is some disagreement over whether syntax is a precise term for these differences. Other terms suggested for these variations in valid R code are 'dialects,' 'interfaces,' and 'domain specific languages.' Throughout this paper, we use the term syntax as a shorthand for these concepts.  At present, there are three primary syntaxes used: base, formula, and `tidyverse` [@mcnamara2018a].

@@

```{r}
library(palmerpenguins)
data("penguins")
penguins <- penguins %>%
  drop_na(bill_length_mm)
library(mosaic)
```

```{r r-syntax, ref = "r-syntax", codecap = "Making a histogram of bill length from the penguins dataset, then taking the mean, using three different R syntaxes. Base syntax is characterized by the dollar sign, formula by the tilde, and tidyvese is dataframe-first. In order for this code to run as-is, missing (NA) values need to be dropped before the code is run.", echo = TRUE, eval = FALSE}
# base syntax
hist(penguins$bill_length_mm)
mean(penguins$bill_length_mm)

# formula syntax
gf_histogram(~bill_length_mm, data = penguins)
mean(~bill_length_mm, data = penguins)

# tidyverse syntax
ggplot(penguins) + 
  geom_histogram(aes(x = bill_length_mm))
penguins %>%
  summarize(mean(bill_length_mm))
```

There is some agreement about pedagogical decisions for teaching R. In particular, most educators agree that in order to reduce cognitive load, instructors should only teach one syntax, and to be as consistent as possible about that syntax [@mcnamaraetal2021a]. There is also some agreement base R syntax is not the appropriate choice for introductory statistics, but there is widespread disagreement on whether the formula syntax or `tidyverse` syntax is better for novices.

@@

Stefik's team is working to add data science functionality to their evidence-based programming language. As a first step toward understanding which elements of existing languages might be best to emulate, they ran an experiment comparing the three main R syntaxes [@rafalskietal2019]. The study showed no statistically significant difference between any of the three syntaxes with regard to time to completion or number of errors. However, there were significant interaction effects between syntax and task, which suggested some syntaxes might be more appropriate for certain tasks [@rafalskietal2019].

Beyond this, examining the results from the study with an eye toward data science pedagogy showed common errors made by students related to their conceptions of dataframes and variables. For example, one of the figures from @rafalskietal2019 shows real student code with errors. In the first line of code, the student gets everything correct using formula syntax, with the exception of the name of the dataframe. When that code does not work, they try again using base R syntax, but again get the dataframe name wrong. After both those failures, they appear to fall back on computer science knowledge and try syntax quite different from R. This is consistent with other studies of novice behavior in R [@roberts2015]. It is not clear if this type of error was dependent on the syntax participants were asked to use.

@@

The remainder of this paper is organized into three sections. Section \ref{sec:methods} describes the setup of the classes, the participants (\ref{sec:participants}) and their experience (\ref{sec:priorexp}), and the content of the course under investigation (\ref{sec:materials}). Section \ref{sec:results} contains results of the analysis, including a comparison of material lengths between the sections (\ref{sec:ll}), the number of unique functions shown in each section (\ref{sec:numfunc}), results from the pre- and post-survey (\ref{sec:prepost}), and analysis of YouTube (\ref{sec:yt}) and RStudio Cloud (\ref{sec:rstudio}) data. Finally, Section \ref{sec:discussion} discusses the results and opportunities for future study.

All pedagogical materials used for the course under discussion are available on GitHub and are Creative Commons licensed, so they can be used or remixed by anyone who wants to use them (\url{https://github.com/AmeliaMN/STAT220-labs}). All code and anonymized data from this paper is also available on GitHub, for reproducibility (\url{https://github.com/AmeliaMN/ComparingSyntaxForModeling}). Data analysis was performed in R, and the paper is written in RMarkdown. The categorical color palette was chosen using Colorgorical [@gramazioetal2017], and colors for the Likert scale plot are from ColorBrewer [@harrowerbrewer2003]. Example data used throughout the paper is from `palmerpenguins` [@horstetal2020]. Code from the formula section uses the R packages loaded in that course, `mosaic` and `ggformula` (`ggformula` is now loaded automatically with `mosaic`) [@pruimetal2017; @kaplanpruim2020]. Code from the tidyverse section uses the functions from that course, the `tidyverse` and `infer` packages  [@wickhametal2019; @brayetal2021].

@@

The author did this head-to-head comparison in her introductory statistics labs. The comparison was run twice, once in the Spring 2020 semester and once in the Fall 2020 semester. The disruption of COVID-19 to the Spring 2020 semester made the resulting data unusable, so this paper focuses on just Fall 2020 data.

@@

@@

Participants were students enrolled in an introductory statistics course at a mid-sized private university in the upper Midwest. At this university, statistics students enroll in a lecture (approximately 60-90 students per section), which is broken into several smaller lab sections for hands-on work in statistical software. Lecture and lab sections are taught by different instructors, and the lab sections associated with a particular lecture often use different software. For example, one lab may use Minitab while the other two use Excel. However, every lab section (no matter what lecture it is associated with, or what software is used) does the same set of standardized assignments. This structure provides a consistent basis for comparison.

In Fall 2020, the author taught two labs associated with the same lecture section, so all students saw the same lecture content. (A third lab was associated with the same lecture, using a different software, and was not considered.) Using random assignment (coin flip), the author selected one lab section to be instructed using formula syntax, and one to be instructed using `tidyverse` syntax. The goal was to compare syntaxes head-to-head.

Because the lab took place during the coronavirus pandemic, the instructor recorded YouTube videos of herself working through the pre-lab documents for each lab, and posted them in advance. Students watched the videos and worked through the associated pre-lab RMarkdown document on their own time, then came to synchronous class to ask questions and get help starting on the real lab assignment. Students used R through the online platform RStudio Cloud [@rstudiopbc2021].

The two labs were of the same size ($n=21$ in both sections) and reasonably similar in terms of student composition. In both sections, approximately half of students were Business majors, with the other half a mix of other majors.

@@

## Prior programming experience {#sec:priorexp}

@@

To verify both groups of students had similar backgrounds, we compared the prior programming experience of the two groups of students. Table \ref{tab:priorexptab} shows results from the pre-survey.  While two additional students in the `tidyverse` section had prior programming experience, the overall pattern was the same. The majority of students in both sections had no prior programming experience.

@@

For the students who had programmed before, none had prior experience with R. Three students had prior experience with Java, 3 with Javascript, and a smaller number had experience with other languages, including C++ and Python.

## Materials {#sec:materials}

Each week, the lab instructor prepared a "pre-lab" document in RMarkdown. The pre-lab covered the topics necessary to complete the standardized lab assignment done by all students across lab sections. Pre-lab documents included text explanations of statistical and R programming concepts, sample code, and blanks (both in the code and the text) for students to fill in as they worked.

The instructor recorded YouTube videos of herself working through the pre-lab documents for each lab, and posted them in advance. Literature about flipped classrooms suggests shorter videos are better for student engagement, although there is no consensus about the ideal length for videos, with suggestions ranging from 5 to 20 minutes as a maximum length for a video [@zuber2016; @beattyetal2019; @guoetal2014]. The instructor attempted to keep the total number of minutes of video content below 20 each week. If video content became too long, the instructor split the content into multiple shorter videos. Students were told to watch the pre-lab video(s) and work through the RMarkdown document on their own time, then come to synchronous class to ask questions and get help starting on the real lab assignment.

The topics covered in Fall 2020 were as follows:

1. [No lab, short week]
2. Describing data: determining the number of observations and variables in a dataset, variable types. 
3. Categorical variables: exploratory data analysis for one or two categorical variables. Frequency tables, relative frequency tables, bar charts, two-way tables, and side-by-side bar charts. 
4. Quantitative variables: exploratory data analysis for one quantitative variable. Histograms, dot plots, density plots, and summary statistics like mean, median, and standard deviation. 
5. Correlation and regression: exploratory data analysis for two quantitative variables. Correlation, scatterplot, simple linear regression as a descriptive technique. 
6. Bootstrap intervals: the use of the bootstrap to construct non-parametric confidence intervals. 
7. Randomization tests: the use of randomization to perform non-parametric hypothesis tests. 
8. Inference for a single proportion: use of the normal distribution to construct confidence intervals and perform hypothesis tests for a single proportion. 
9. Inference for a single mean: use of the t-distribution to construct confidence intervals and perform hypothesis tests for a single mean. 
10. Inference for two samples: use of distributional approximations (normal or t) to perform inference for a difference of proportions or a difference of means.
11. [No lab, assessment]
12. [No lab, Thanksgiving]
13. ANOVA: inference for more than two means, using the F distribution. 
14. Chi-square: inference for more than two counts, using the $\chi^2$ distribution
15. Inference for Regression: inference for the slope coefficient in simple linear regression, prediction and confidence intervals. Multiple regression.

Although this was a 15-week semester, there are only 12 lab topics. Labs were not held during the first week of classes or during Thanksgiving week. Additionally, there were two "lab assessments" to gauge student understanding of concepts within the context of their lab software. One took place during finals week, the other was scheduled in week 11.

One obvious question arising when considering the comparison of the two syntaxes is whether students performed better in one section or another. The IRB did not cover examining student work (an obvious place for improved further research), so we cannot look at student outcomes on a per-assignment basis. However, running a randomization test for a difference in overall mean lab grades showed no significant difference between the two sections. While they may have been interesting differences in grades depending on the topic of the lab, we at least know these differences averaged out in the end.

Similarly, it would be interesting to know if student attitudes about the instructor were different from the summative student evaluations completed by all students at the end of the semester. These evaluations are anonymous, and the interface only provides summary statistics. Again, a test for a difference in means showed no difference in mean evaluation score on the questions "Overall, I rate this instructor an excellent teacher." and "Overall, I rate this course as excellent."

## Lab lengths {#sec:ll}

The first question we seek to answer is whether the materials presented to students were of approximately the same length. We can assess this based on the length of the pre-lab documents (in lines) and of the pre-lab videos (in minutes).

The length of the pre-lab RMarkdown documents can be measured using lines. Figure \ref{fig:prelablength} shows the number of lines of code for each section's pre-lab document, per week. Lines in the RMarkdown document include the YAML header (consistent between documents), the descriptive text about processes (largely similar between documents), and the code in code chunks, which varies based on the syntax of the lab.

@@

```{r}
labsfortesting <- lablines %>%
  drop_na(file) %>%
  pivot_wider(week, names_from = section, values_from = lines) %>%
  mutate(diff = tidyverse - formula)
```

```{r, eval = FALSE}
qqnorm(labsfortesting$diff)
qqline(labsfortesting$diff)
```

```{r}
labttest <- tidy(t.test(labsfortesting$diff))
library(infer)
set.seed(11082021)
labboots <- labsfortesting %>%
  specify(response = diff) %>%
  generate(reps = 5000, type = "bootstrap") %>%
  calculate(stat = "mean")

labboot <- labboots %>%
  get_ci()
```

```{r, eval = FALSE}
qqnorm(labboots$stat)
qqline(labboots$stat)
```

```{r videolength, fig.cap = "Length of pre-lab videos each week. Outlines help delineate multiple videos for a single week."}
youtube_videos <- youtube_videos %>%
  mutate(week_topic = as_factor(week_topic)) %>%
  mutate(week_topic, fct_reorder(week_topic, week)) %>%
  complete(week = full_seq(1:15, 1), nesting(section))

@@

```{r}
youtube_fortesting <- youtube_weeks %>%
  select(week, section, tot_min) %>%
  pivot_wider(week, names_from = section, values_from = tot_min) %>%
  mutate(diff = tidyverse - formula, 
         avglength = (tidyverse + formula)/2, 
         perc_diff = abs(tidyverse - formula)/avglength)
```

```{r, eval = FALSE}
qqnorm(youtube_fortesting$diff)
qqline(youtube_fortesting$diff)
```

```{r}
yttest <- tidy(t.test(youtube_fortesting$diff))
ytboots <- youtube_fortesting %>%
  specify(response = diff) %>%
  generate(reps = 5000, type = "bootstrap") %>%
  calculate(stat = "mean") 
ytboot <- ytboots %>%
  get_ci(point_estimate = mean(youtube_fortesting$diff), type = "bias-corrected")
detach("package:infer")
```

```{r, eval = FALSE}
qqnorm(ytboots$stat)
qqline(ytboots$stat)
```

```{r, eval = FALSE}
linesvminutes <- lablines %>%
  left_join(youtube_videos, by = c("week", "section"))
m1 <- lm(tot_min~lines, data = linesvminutes)
m2 <- lm(tot_min~lines+section, data = linesvminutes)
m3 <- lm(tot_min~lines*section, data = linesvminutes)
anova(m1, m2, m3)
summary(m1)

ggplot(linesvminutes, aes(x=lines, y=tot_min, color = section)) + geom_point() 
```

Every attempt was made to align these RMarkdown documents, so the descriptive text was only changed when necessary to describe specific elements of the code. Similarly, if blank code chunks appeared in one lab, that was mirrored by a blank chunk in the other lab. Both labs' documents were styled using the styler package [@mullerwalthert2022] to remove inconsistencies with spacing, assignment operators, and the like. Code was styled using the default style from the package, which is based on the tidyverse style guide [@wickham2022]. The tidyverse style guide has become the de facto style guide for R, as previously-existing style guides like the Google R Style Guide have largely decided to follow @wickham2022.

Figure \ref{fig:prelablength} indicates RMarkdown documents for the `tidyverse` section tended to be longer. We can compute a difference in lab lengths for each week, and compute the mean difference, which is `r round(mean(labsfortesting$diff))` lines. Although we only have 12 labs worth of data, it appeared relatively normally distributed. A 95\% confidence interval for the difference in line length (computed using the Student's t-distribution) is (`r round(labttest$conf.low)`, `r round(labttest$conf.high)`), and a bootstrap confidence interval computed with 5,000 bootstrap samples and using the percentile method is (`r round(labboot$lower_ci)`, `r round(labboot$upper_ci)`). Both intervals indicate labs for the `tidyverse` section were longer, but only by a few lines.

A slightly longer length for these labs makes sense, because `tidyverse` code is characterized by multiple short lines strung together into a pipeline with `%>%`, while the formula syntax typically has single function calls, sometimes with more arguments. The code shown in this paper is styled the same way the labs were, so any code comparison you see (for example, in \ref{tally-ex1} and \ref{tidy-tally1}) will show the difference in lines of code between the two syntaxes.

Then the question becomes if the longer lengths of documents lent themselves to longer pre-lab videos. Figure \ref{fig:videolength} shows the video lengths, which appear more consistent between sections. Effort was made to ensure the maximum video length was approximately 20 minutes, and some weeks had multiple videos.

Again, we can compute a pairwise difference in total video length (adding together multiple videos in weeks that had them), and compute the mean of that difference. That difference is `r round(mean(youtube_fortesting$diff))` minutes (`tidyverse` videos being longer). The distribution of differences appeared to be right-skewed, which may impact results because the sample size is so low. Using a t-distribution, we can compute a 95\% confidence interval, which is (`r round(yttest$conf.low, 1)`, `r round(yttest$conf.high, 1)`), and thus contains 0.

>>>>

abstract: >
  This paper reports on a head-to-head comparison run in a pair of introductory statistics labs, one conducted fully in the formula syntax, the other in tidyverse. Analysis of pre- and post-survey data show no differences between the two labs, so students appeared to have a positive experience regardless of section. Analysis of incidental data from YouTube and RStudio Cloud show interesting distinctions. The formula section appeared to watch a larger proportion of pre-lab YouTube videos, but spend less time computing on RStudio Cloud. Conversely, the tidyverse section watched a smaller proportion of the videos and spent more time computing. Analysis of lab materials showed that tidyverse labs tended to be slightly longer in terms of lines in the provided RMarkdown materials and minutes of the associated YouTube videos. Both labs relied on a quite small vocabulary of consistent functions, which can provide a starting point for instructors interested in teaching introductory statistics in R. The tidyverse labs exposed students to slightly more distinct R functions, but reused functions more frequently. This work provides additional evidence for instructors looking to choose between syntaxes for introductory statistics teaching. 
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    keep_tex: yes
    includes:
      in_header: header.tex
    citation_package: natbib
header-includes:
-  \usepackage{amsmath}
editor_options: 
  chunk_output_type: console
---

\doublespacing

@@

```{r, cache = FALSE}
library(here)
library(tidyverse)
library(kableExtra)
library(lme4)
library(broom.mixed)
library(bookdown)
library(ggh4x)
library(cowplot)
ggplot2::theme_set(ggplot2::theme_gray(base_size = 14))
```

@@

For example, the UC Berkeley Data 8 course uses Python, including elements of the commonly-used `matplotlib` and `numpy` libraries as well as a specialized library written to accompany the curriculum called `datascience` [@adhikarietal2021; @deneroetal2022]. The `datascience` library was designed to reduce complexity in the code. At the K-12 level, the language `Pyret` has been developed as a simplified version of Python to accompany the Bootstrap Data Science curriculum [@krishnamurthietal2020].

In R, the development of less-complex code for students has been under consideration for even longer [@pruimetal2011]. R offers non-standard evaluation, which allows package authors to create new 'syntax' for their packages [@morandatetal2012]. In human language, syntax is the set of rules for how words and sentences should be structured. If you use the wrong syntax in human language, people will hear there is something wrong with your speech or writing. However, because human understanding is flexible, the listener will probably still understand the general idea you were trying to convey. Syntax in programming language is more formal-- it governs what code will execute, run, or compile correctly. Using the wrong syntax usually means failing to get a result from the program.

Typically, programming languages have only one valid syntax. For example, an aphorism about Python is "There should be one-- and preferably only one --obvious way to do it" [@peters2004]. But, non-standard evaluation in R has allowed there to be many obvious ways to do the same task. There is some disagreement over whether syntax is a precise term for these differences. Other terms suggested for these variations in valid R code are 'dialects,' 'interfaces,' and 'domain specific languages.' Throughout this paper, we use the term syntax as a shorthand for these concepts. At present, there are three primary syntaxes used: base, formula, and `tidyverse` [@mcnamara2018a].

@@

```{r, eval = FALSE}
library(palmerpenguins)
data("penguins")
# penguins <- penguins %>%
#   drop_na(bill_length_mm)
library(mosaic)
library(tidyverse)
```

```{r r-syntax, ref = "r-syntax", codecap = "Making a histogram of bill length from the penguins dataset, then taking the mean, using three different R syntaxes. Base syntax is characterized by the dollar sign, formula by the tilde, and tidyvese is dataframe-first.", echo = TRUE, eval = FALSE}
# base syntax
hist(penguins$bill_length_mm)
mean(penguins$bill_length_mm, na.rm = TRUE)

# formula syntax
gf_histogram(~bill_length_mm, data = penguins)
mean(~bill_length_mm, data = penguins, na.rm = TRUE)

# tidyverse syntax
ggplot(penguins) + 
  geom_histogram(aes(x = bill_length_mm))
penguins %>%
  drop_na(bill_length_mm) %>%
  summarize(mean(bill_length_mm))
```

There is some agreement about pedagogical best practices while teaching R. In particular, most educators agree that in order to reduce cognitive load, instructors should only teach one syntax, and should be as consistent as possible about that syntax [@mcnamaraetal2021a]. There is also some agreement that base R syntax is not the appropriate choice for introductory statistics, but there is widespread disagreement on whether the formula syntax or `tidyverse` syntax should be used in these courses.

@@

Rafalski et al (a group which includes Stefik) ran an experiment comparing the three main R syntaxes. The study showed no statistically significant difference between the three syntaxes with regard to time to completion or number of errors. However, there were significant interaction effects between syntax and task, which suggested some syntaxes might be more appropriate for certain tasks [@rafalskietal2019].

Examining the results from the study with an eye toward data science pedagogy showed common errors made by students related to their conceptions of dataframes and variables. For example, one of the figures from @rafalskietal2019 shows real student code with errors. In the first line of code, the student gets everything correct using formula syntax, with the exception of the name of the dataframe. When that code does not work, they try again using base R syntax, but again get the dataframe name wrong. After both those failures, they appear to fall back on computer science knowledge and try syntax quite different from R. This is consistent with other studies of novice behavior in R [@roberts2015]. It is not clear if this type of error was dependent on the syntax participants were asked to use.

@@

In this paper, we seek to more deeply compare the experience of teaching and learning R in introductory statistics classes using the formula and `tidyverse` syntaxes. If you are looking for a definitive answer on which syntax is 'best,' you will be disappointed with the conclusions of this paper. But, if you are an instructor trying to weigh the benefits and drawbacks of each syntax, this paper will provide additional questions and evidence you can use to guide your decision.

The remainder of this paper is organized into four sections. Section \ref{sec:methods} describes the setup of the classes used in the comparison (\ref{sec:structure}), the participants (\ref{sec:participants}), and the content of the course under investigation (\ref{sec:materials}), including the RMarkdown documents provided to students (\ref{sec:rmd}) and the pre-lab YouTube videos for the flipped class (\ref{sec:videolengths}). Section \ref{sec:results} contains results of the pilot study, including a discussion about the lack of summative assessments (\ref{sec:assessment}), results from the pre- and post-survey (\ref{sec:prepost}), analysis of YouTube (\ref{sec:yt}) and RStudio Cloud (\ref{sec:rstudio}) data, a study of the functions shown in each section (\ref{sec:numfunc}), and a qualitative assessment of divergent labs (\ref{sec:diflabs}). Finally, Section \ref{sec:discussion} discusses the results and opportunities for future study.

@@

The author conducted a head-to-head comparison of the formula and `tidyverse` syntaxes in her introductory statistics labs. The comparison was run twice, once in the Spring 2020 semester and once in the Fall 2020 semester. The disruption of COVID-19 to the Spring 2020 semester made the resulting data unusable, so this paper focuses on just Fall 2020 data.

@@

<!-- REDACTED IRB number -->

## Course structure {#sec:structure}

This comparison was run in two introductory statistics labs at a mid-sized private university in the upper Midwest. At this university, statistics students enroll in a lecture (approximately 60-90 students per section), which is broken into several smaller lab sections for hands-on work in statistical software. Lecture and lab sections are taught by different instructors, and the lab sections associated with a particular lecture section often use different software. For example, a lecture section of 90 students may have one lab section of 30 students using Minitab while the other two sections use Excel. However, every lab section (no matter what lecture it is associated with, or what software is used) does the same set of standardized assignments. This structure provides a consistent basis for comparison.

In Fall 2020, the author taught two labs sections associated with the same lecture section, meaning all students saw the same lecture content. (A third lab section was associated with the same lecture, using a different software, and was not considered.) Using random assignment (coin flip), the author selected one lab section to be instructed using formula syntax, and one to be instructed using `tidyverse` syntax. The goal was to compare syntaxes head-to-head.

The lab took place during the coronavirus pandemic, and the lab instructor used a flipped classroom format. Each week, the instructor prepared a "pre-lab" document in RMarkdown. The pre-lab covered the topics necessary to complete the standardized lab assignment done by all students across lab sections. Pre-lab documents included text explanations of statistical and R programming concepts, sample code, and blanks (both in the code and the text) for students to fill in as they worked. Students in both sections used R through the online platform RStudio Cloud [@rstudiopbc2021].

The instructor recorded YouTube videos of herself working through the pre-lab documents for each lab, and posted them in advance. Students were expected to watch the pre-lab video(s) and work through the RMarkdown document on their own time, then come to synchronous class to ask questions and get help starting on the lab assignment.

Literature about flipped classrooms suggests shorter videos are better for student engagement, although there is no consensus about the ideal length for videos, with suggestions ranging from 5 to 20 minutes as a maximum length for a video [@zuber2016; @beattyetal2019; @guoetal2014]. The instructor attempted to keep the total number of minutes of video content below 20 each week. If video content became too long, the instructor split the content into multiple shorter videos.

@@

Participants were students enrolled in two introductory statistics lab sections associated with a common introductory statistics course. The two labs were of the same size ($n=21$ in both sections) and reasonably similar in terms of student composition. In both sections, approximately half of students were Business majors, with the other half a mix of other majors.

@@

@@

The course these labs were associated with is an introductory statistics course, so it is a first introduction to statistics for all but a handful of students. Typically, students also have no prior experience programming. To verify this, we compared the prior programming experience of the two groups of students. Table \ref{tab:priorexptab} shows results from the pre-survey. While two additional students in the `tidyverse` section had prior programming experience, the majority of students in both sections had no prior programming experience.

@@

For the students who had programmed before, none had prior experience with R. Three students had prior experience with Java, three with Javascript, and a smaller number had experience with other languages, including C++ and Python.

## Topics and materials {#sec:materials}

The course these labs accompanied was an introductory statistics course using the Lock5 textbook [@locketal2020]. Weekly lab topics were aligned with lecture content, and standardized across all labs for the semester (13 lab sections using a variety of different software). The topics covered in Fall 2020 are shown in Table \ref{topics}.

```{=tex}
\begin{table}
\singlespacing
\begin{tabular}{p{\linewidth}}
1. [No lab, short week] \\ 
2. Describing data: determining the number of observations and variables in a dataset, variable types. \\
3. Categorical variables: exploratory data analysis for one or two categorical variables. Frequency tables, relative frequency tables, bar charts, two-way tables, and side-by-side bar charts. \\
4. Quantitative variables: exploratory data analysis for one quantitative variable. Histograms, dot plots, density plots, and summary statistics like mean, median, and standard deviation. \\
5. Correlation and regression: exploratory data analysis for two quantitative variables. Correlation, scatterplot, simple linear regression as a descriptive technique. \\
6. Bootstrap intervals: the use of the bootstrap to construct non-parametric confidence intervals. \\
7. Randomization tests: the use of randomization to perform non-parametric hypothesis tests. \\
8. Inference for a single proportion: use of the normal distribution to construct confidence intervals and perform hypothesis tests for a single proportion. \\
9. Inference for a single mean: use of the t-distribution to construct confidence intervals and perform hypothesis tests for a single mean. \\
10. Inference for two samples: use of distributional approximations (normal or t) to perform inference for a difference of proportions or a difference of means.\\
11. [No lab, assessment]\\
12. [No lab, Thanksgiving]\\
13. ANOVA: inference for more than two means, using the F distribution.\\
14. Chi-square: inference for more than two counts, using the $\chi^2$ distribution.\\
15. Inference for Regression: inference for the slope coefficient in simple linear regression, prediction and confidence intervals. Multiple regression. \\
Finals. [Second assessment] \\
\end{tabular}
\caption{List of topics for labs across the 15-week semester.}\label{topics}
\end{table}
\doublespacing
```
Although this was a 15-week semester, there are only 12 lab topics. Labs were not held during the first week of classes or during Thanksgiving week. Additionally, there were two "lab assessments" to gauge student understanding of concepts within the context of their lab software. One took place during finals week, the other was scheduled in week 11.

### RMarkdown documents {#sec:rmd}

The instructor prepared weekly RMarkdown documents for students, as described in Section \ref{sec:structure}. One question is whether the materials presented to students were of approximately the same length. We can first assess this using the length of the pre-lab RMarkdown documents, measured using lines. Figure \ref{fig:prelablength} shows the number of lines of code and text in each section's pre-lab document, per week. Lines in the RMarkdown document include the YAML header (consistent between documents), the descriptive text about processes (largely similar between documents), and the code in code chunks, which varied based on the syntax of the lab.

@@

Every attempt was made to align these RMarkdown documents, so the descriptive text was only changed when necessary to describe specific elements of the code. Similarly, if blank code chunks appeared in one lab, that was mirrored by a blank chunk in the other lab. Both labs' documents were styled using the `styler` package [@mullerwalthert2022] to remove inconsistencies with spacing, assignment operators, and the like. The `styler` package is based on the tidyverse style guide [@wickham2022], and has become the de facto style guide for R. Previously-existing style guides like the Google R Style Guide have largely decided to follow @wickham2022.

Figure \ref{fig:prelablength} indicates RMarkdown documents for the `tidyverse` section tended to be longer. A slightly longer length for `tidyverse` materials makes sense, because `tidyverse` code is characterized by multiple short lines strung together into a pipeline with `%>%`, while the formula syntax typically has single function calls, sometimes with more arguments. The code shown in this paper is styled the same way as the labs, so any code comparison you see (for example, in \ref{r-syntax} or \ref{tally-ex1} and \ref{tidy-tally1}) will show the difference in lines of code between the two syntaxes.

```{r}
labsfortesting <- lablines %>%
  drop_na(file) %>%
  pivot_wider(week, names_from = section, values_from = lines) %>%
  mutate(diff = tidyverse - formula,
         perc_diff = (diff/formula)*100) %>%
  drop_na(week)
```

We can quantify how much longer the `tidyverse` documents were, either in terms of absolute lines (`tidyverse` labs were on average `r round(mean(labsfortesting$diff))` lines longer) or in terms of percent difference (`tidyverse` labs were, on average, `r round(mean(labsfortesting$perc_diff))`% longer).

```{r videolength, fig.cap = "Length of pre-lab videos each week. Outlines delineate multiple videos for a single week."}
youtube_videos <- youtube_videos %>%
  mutate(week_topic = as_factor(week_topic)) %>%
  mutate(week_topic, fct_reorder(week_topic, week)) %>%
  complete(week = full_seq(1:15, 1), nesting(section))

@@

### YouTube vidoes {#sec:videolengths}

Another major instructional source were the YouTube videos the instructor recorded of herself working through the pre-lab documents for each lab.

Figure \ref{fig:videolength} shows the lengths of videos over the course of the semester. Video length appears generally consistent between sections. Effort was made to ensure the maximum video length was approximately 20 minutes, although some weeks had multiple videos.

```{r}
youtube_fortesting <- youtube_weeks %>%
  select(week, section, tot_min) %>%
  pivot_wider(week, names_from = section, values_from = tot_min) %>%
  mutate(diff = tidyverse - formula, 
         avglength = (tidyverse + formula)/2, 
         perc_diff = (diff/formula)*100)
```

We can again compute the percentage difference in total video length (adding together multiple videos in weeks that had them), and compute the mean of that difference. This shows us that `tidyverse` labs were `r round(mean(youtube_fortesting$diff))` minutes longer or `r round(mean(youtube_fortesting$perc_diff))`% longer than formula videos, on average. The `tidyverse` labs appear to be slightly longer in of lines of code and video length, although these differences are slight.

The results from this work can be split into quantitative and qualitative evidence. Quantitative results can be derived from the pre- and post-survey in the class (the most intentionally-collected data), as well as YouTube Analytics and RStudio Cloud usage data (incidental data). We can also quantify the number of functions used in both sections by analyzing the RMarkdown documents as data in their own right. All of these quantitative results show slight differences between sections, which hint toward some of the qualitative differences in experience. We also discuss these qualitative differences by highlighting three topics where the experience teaching and learning the two syntaxes diverged.

One obvious question arising when considering the comparison of the two syntaxes is whether students performed better in one section or another. The IRB did not cover examining student work (an obvious place for improved further research), so we cannot look at student outcomes on a per-assignment basis. However, running a randomization test for a difference in overall mean lab grades showed no significant difference between the two sections. These grades were raw scores, and were not curved in any way. Lab grades comprise 30% of overall lecture course grade. While there may have been interesting differences in grades depending on the topic of the lab, we at least know these differences averaged out in the end.

Similarly, it would be interesting to know if student attitudes about the instructor were different from the summative student evaluations completed by all students at the end of the semester. These evaluations are anonymous, and the interface only provides summary statistics. Again, a randomization test for a difference in means showed no difference in mean evaluation score on the questions "Overall, I rate this instructor an excellent teacher." and "Overall, I rate this course as excellent."

While we cannot make conclusions about summative assessments, there are a number of other data sources to give insight into the difference between the experience of teaching and learning the two syntaxes.

Students in both sections were given a pre- and post-survey on their experience in the class. The majority of the survey was modeled on a pre- and post-survey used by The Carpentries, a global nonprofit teaching coding skills [@thecarpentries2021].

As discussed in Section \ref{sec:participants}, the number of students who completed both the pre- and post-surveys was low ($n=8$ for the formula section, and $n=13$ for the `tidyverse` section), so there are major limitations to paired analysis.

On the survey, respondents were asked to use a 5-step Likert scale, from 1 (strongly disagree) to 5 (strongly agree) to rate their agreement with the following statements:

````{=tex}
\begin{itemize}
\itemsep-3mm
```{r, results = 'asis'}
qs %>%
  select(Q11.x:Q16.x) %>%
  pivot_longer(cols = everything(), names_to = "old", values_to = "q") %>%
  select(-old) %>%
  mutate(q = paste("\\item", q, "\n")) %>%
  pull(q) %>%
  walk(cat)
```
\end{itemize}
````

In Figure \ref{fig:pre-post}, you can see a visualization of these Likert-scale questions.The visualization shows the pre- and post-survey responses to the survey questions, broken down by section. There is no obvious overall trend. However, considerable improvement was seen in in the Programming Confident category, which suggests that students in both sections became more confident in their "ability to make use of programming software to work with data."

Other questions showed small gains or unclear changes, like the Analyses Easier category. In the formula section, this question appears to show a small increase in agreement from pre- to post-survey, while the `tidyverse` section shows either no change or perhaps a slight improvement (more students saying they strongly agree, but fewer in the positive answers overall). Other questions show a decrease in agreement from the pre- to the post-survey, like Overcome Problem. Finally, some questions show improvement in one section but a decline in agreement in the other, like Search Online. In the `tidyverse` section, there appears to be an increase in their confidence in searching "for answers to [their] technical questions online," whereas in the formula section there is a decrease.

```{r}
pre_sum <- prepostsurvey %>%
  select(Section.x, Q11.x:Q16.x) %>%
  drop_na(Section.x) %>%
  pivot_longer(-Section.x,
    names_to = "skill",
    values_to = "feeling"
  ) %>%
  rename(Section = Section.x)
